Classless Java

### 2015-11-15 paper/sections/Formalization.tex
revised method overriding mechanism in Java
Some work on the table for java bug

\subsection{Method Overriding Mechanism}
We summarize the mechanism for method overriding below (all methods considered
here are methods with the same name and argument types):

1. Two abstract methods with subtyping relation do not cause conflict, and the
most specific one wins. Subtyping relation for two methods means they have the
same arguments' types and one return type is a subtype of the other one. For
example:
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}
Method \texttt{Integer m()} in \texttt{B} wins over method \texttt{Object m()}
in \texttt{A} because \texttt{Integer <: Object}.

2. A default method conflict with another default method or abstract method. For
example:
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}

3. The method in the current interface wins over any methods defined in its
super-interfaces, provided that the method conform to the subtype of all methods
in its super-interfaces, i.e., the method is the most specific one. For example:
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { int m(); } //accepted
\end{lstlisting}

\subsection{Method Overriding Bug in Java}
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & ERROR                        & ERROR       & ERROR       \\ \hline
\textbf{B2}        & A1/B2                        & ERROR       & A3/B2       \\ \hline
\textbf{B3}        & \cellcolor[HTML]{C0C0C0}B3.m & ERROR       & ERROR       \\ \hline
\end{tabular}
\caption{M extends $A_i, B_j$}
\label{table:javabug}
\end{table}

In our approach, we overcome this bug by detecting method correctness when
typechecking all interfaces.

@@

As you can see, we are delegating the work to three others auxiliary functions: $\tops(\Cs), \shadow(\m,\Cs) and \override(\method,\method')$
${}_{}$\\*
\marco{we need to discuss the names tops, shadow and override}

@@

\method \subtype
\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}=
\method\subtype\mh\QM;
\\
\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\subtype\method=
\mh\QM;\subtype\method
\end{array}$
${}_{}$\\*
Where $\Aux{mostSpecific}$ return the most specific method using return type specialization as introduced in Java ??
We just check the subtype between method headers, so we discard information abut method implementation.\\*
${}_{}$\\*
The override function models how the implementation in an interface can override implementation in the superinterface; even in case of a conflict.
Note how we use the special value $\none$, and how (forth case) overriding can solve a conflict.
\\*
\!\!\!\!$\begin{array}{ll}
\override(\none,\none)=\none\\
\override(\method,\none)=\method\\

\override(\none,\method)=\method
\whereNote
\mwhere\ \method\mbox{ not of the form } \conflicted\ \mh\QM;\\
\override(\method,\method')
=
\method
\whereNote
\mwhere\ \method'\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'
\\
%\override(\method,\mh')
%=\method
%\whereNote
%\mif\ \method\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@} \}, \mh\subtype\mh'
%\override(\_,\_)=\error & Otherwise
%\textsf{shadow}(body_1, body_2)=\emptyset & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_1 & \textsf{if }body_2=\emptyset\textsf{ or }body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_2 \hspace{.1in}  & \textsf{if }body_1=\emptyset\textsf{ or }body_1.\textsf{modifier}=\textbf{static}
\end{array}$

@@

>>>>

@@

As you can see, we are delegating the work to three others auxiliary functions:\\*
 $\tops(\Cs), \shadow(\m,\Cs)$ and $\override(\method,\method')$
${}_{}$\\*
\marco{we need to discuss the names tops, shadow and override}

@@

\method \subtype
\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}=
\method\subtype\mh\QM;
\\
\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}\subtype\method=
\mh\QM;\subtype\method
\end{array}$
${}_{}$\\*
Where $\Aux{mostSpecific}$ return the most specific method using return type specialization as introduced in Java\marco{insert the version number}
We just check the subtype between method headers, so we discard information abut method implementation.\\*
${}_{}$\\*
The override function models how the implementation in an interface can override implementation in the superinterface; even in case of a conflict.
Note how we use the special value $\none$, and how (forth case) overriding can solve a conflict.
\\*
\!\!\!\!$\begin{array}{ll}
\override(\none,\none)=\none\\
\override(\method,\none)=\method\\

\override(\none,\method)=\method
\whereNote
\mwhere\ \method\mbox{ not of the form } \conflicted\ \mh\QM;\\
\override(\method,\method')
=
\method
\whereNote
\mwhere\ \method'\in\{\mh\QM;,\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'
\\
%\override(\method,\mh')
%=\method
%\whereNote
%\mif\ \method\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@} \}, \mh\subtype\mh'
%\override(\_,\_)=\error & Otherwise
%\textsf{shadow}(body_1, body_2)=\emptyset & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_1 & \textsf{if }body_2=\emptyset\textsf{ or }body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_2 \hspace{.1in}  & \textsf{if }body_1=\emptyset\textsf{ or }body_1.\textsf{modifier}=\textbf{static}
\end{array}$

@@

\subsection{Method Overriding Mechanism}
We summarize the mechanism for method overriding below (all methods considered
here are methods with the same name and argument types):
\begin{enumerate}
\item Two abstract methods with subtyping relation do not cause conflict, and the
most specific one wins. Subtyping relation for two methods means they have the
same arguments' types and one return type is a subtype of the other one. For
example:
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}
Method \texttt{Integer m()} in \texttt{B} wins over method \texttt{Object m()}
in \texttt{A} because \texttt{Integer <: Object}.

\item A default method conflict with another default method or abstract method. For
example:
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is in contrast with what happens in most trait models, where \Q@D2@ would be accepted, and the implementation in \Q@A2@ would be part of the behaviour of \Q@D2@.

\item The method in the current interface wins over any methods defined in its
super-interfaces, provided that the method conform to the subtype of all methods
in its super-interfaces, i.e., the method is the most specific one. For example:
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { int m(); } //accepted
\end{lstlisting}
\end{enumerate}
\subsection{Method Overriding Incoherence in Java}
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
While trying to formally encode the Java specification we have done some tests to clarify corner case behaviour.
Consider the following correct declarations:
\begin{lstlisting}
interface A1{T m(); }
interface A2 extends A1{default T m(){ ... } }
interface A3 extends A2{T m(); }

interface B1{default T m(){ ... } }
interface B2 extends B1{T m(); }
interface B3 extends B2{default T m(){ ... } }
\end{lstlisting}

What happens if we define a new interface \Q@M@ implementing one \Q@A@${}_i$ and one \Q@B@${}_i$? we have 9 cases, that can fit nicely a table:\\*

\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & error                        & error      &error       \\ \hline
\textbf{B2}        & accepted                        & error       & accepted       \\ \hline
\textbf{B3}        & \cellcolor[HTML]{C0C0C0}accepted &error       & error      \\ \hline
\end{tabular}
\\*
 and we have found that javac\footnote{We refer to \texttt{javac} as in JDK version is\marco{add}}
have a very strange behaviour in the following case

In our approach, we overcome this bug by detecting method correctness when
typechecking all interfaces. Internally, algorithm in the auxiliary function
$\mBody(\m,\C)$ apply the correct method overriding mechanism and returns the
correct method signature.

### 2015-11-15 paper/sections/Overview.tex
Pegasus vector-graph
Some work on the table for java bug

\begin{figure}[tbp]
\centering
\includegraphics[width=4in]{PegasusDetail.png}
\caption{Pegasus Example.}\label{fig:pegasus}
\end{figure}

@@

%Suppose we want to model an animal system as shown in Fig~\ref{fig:pegasus}. In
%order to model \texttt{Pegasus}, \emph{multiple inheritance} is clearly needed
%in this example. Therefore, in Java we use interfaces instead of classes. An
%\texttt{Animal} has a \texttt{point} field, indicating the current position of
%the animal. \texttt{Horse} and \texttt{Bird} are subtypes of \texttt{Animal},
%with methods \texttt{run()} and \texttt{fly()}, respectively. Note that
%\texttt{Bird} can fly, therefore refines the type of \texttt{point} from
%\texttt{Point2D} to \texttt{Point3D}. Pegasus (one of the best known creatures
%in Greek mythology) can not only \emph{run} but also \emph{fly}! This is the
%place where \emph{``multiple inheritance''} is needed, because \texttt{Pegasus}
%need to obtain \texttt{fly} and \texttt{run} functionality from both
%\texttt{Horse} and \texttt{Bird}.
%---------------
% We can put more in dept discussion while discussing the classes in details.
%--------------
\subsubsection{\Q@Point2D@: simple immutable data with fields}
Since in ClassLess Java there are no classes, our \Q@Point2D@ must be an interface, as for example:
\begin{lstlisting}
interface Point2D{ int x(); int y();}
\end{lstlisting}
That is, since Java disallows fields inside interfaces,
 we simulate state by abstract methods.
However, how can we create an instance of \Q@Point2D@ without using classes? 
we can just provide an implementation for \Q@x()@ and \Q@y()@ as in
\begin{lstlisting}
new Point2D(){ public int x(){return 4;} public int y(){return 2;}}
\end{lstlisting}
However we are not required to use this cumbersome syntax\footnote{Available in Java from version ...} for every object allocation.
As programmes do, for ease or reuse, we can encapsulate boring long repetitive code in a method. A static factory method in this case is appropriate:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
  }
\end{lstlisting}
This obvious constructor code can be automatically generated by our \mixin annotation.

@@

More precisely, a field or factory parameter is generated for every no-args method requiring implementation whose name does not have special meaning\footnote{Formal definition of \mixin behaviour is provided later.}.
An example of code using \Q@Point2D@ is 
\Q@Point2D.of(42,myPoint.y())@
where we return a new point, using \Q@42@ as x-coordinate, and 
taking all the other informations (only \Q@y@ in this case) from another point.
This pattern is very common when programming with immutable data-structures; it is so common that we decided to support it in our code generation as \Q@with-@ methods, that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x(); int y();
  Point2D withX(int val); 
  Point2D withY(int val);}
\end{lstlisting}
is equivalent to 
\begin{lstlisting}
@Mixin interface Point2D{ int x(); int y();
  default Point2D withX(int val){ return Point2D.of(val,this.y());}
  default Point2D withY(int val){ return Point2D.of(this.x(),val);}  }
\end{lstlisting}
Notice how we can provide implementation for methods in an interface, 
and our annotation generate code only if the method \emph{needs} implementation. In this way is trivial for the programmer to personalize the behaviour if they have any special need.

@@

Just to give you a feeling on how much boring code \mixin is generating,  we show the correspondent code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);
  Point3D withY(int val);
  Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int x, int y, int z){return new Point3D(){
    public int x(){return x;} 
    public int y(){return y;}
    public int z(){return z;}
    public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
    public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
    public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
    public Point3D with(Point2D val){
      if(val instanceof Point3D){return (Point3D)val;}
      return Point3D.of(val.x(),val.y(),this.z());
      }
    };
  }
}
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

\Q@Bird@s are \Q@Animal@s, 
but while \Q@Animal@s, only need 2D locations,
\Q@Bird@s need  3D locations.
In Java if we define an animal class with a field we have a set of unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@:
this is bad since all animals would require more that is needed, and also it may requires the programmer to predict the future, or it may require to adapt the old code to accommodate for new evolutions.
\item Define a \Q@Point2D@ field in \Q@Animal@ and...
\item Define an extra \Q@int z@ field in \Q@Bird@.
This solution is very at-hoc, requires to basically duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside of \mbox{\Q@Bird@}.
Again, there are many reasons this would be bad, the most dramatic is that it would not scale to a scenario when the programmer of \Q@Bird@ and the programmer of \Q@Point3D@ are different.
\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects in the field and cast
the value out of the \Q@Point2D@ field to \Q@Point3D@ when implementing the overridden getter.
This solution scales to the multiple programmers approach, but requires ugly casts and can be implemented in a wrong way leading to bugs.
\end{itemize}

>>>>

\begin{figure}
\centering
\includegraphics[height=4cm]{PegasusDetail.pdf}
\caption{Pegasus Example.}\label{fig:pegasus}
\end{figure}

@@

%Suppose we want to model an animal system as shown in Fig~\ref{fig:pegasus}. In
%order to model \texttt{Pegasus}, \emph{multiple inheritance} is clearly needed
%in this example. Therefore, in Java we use interfaces instead of classes. An
%\texttt{Animal} has a \texttt{point} field, indicating the current position of
%the animal. \texttt{Horse} and \texttt{Bird} are subtypes of \texttt{Animal},
%with methods \texttt{run()} and \texttt{fly()}, respectively. Note that
%\texttt{Bird} can fly, therefore refines the type of \texttt{point} from
%\texttt{Point2D} to \texttt{Point3D}. Pegasus (one of the best known creatures
%in Greek mythology) can not only \emph{run} but also \emph{fly}! This is the
%place where \emph{``multiple inheritance''} is needed, because \texttt{Pegasus}
%need to obtain \texttt{fly} and \texttt{run} functionality from both
%\texttt{Horse} and \texttt{Bird}.
%---------------
% We can put more in dept discussion while discussing the classes in details.
%--------------
\subsubsection{\Q@Point2D@: simple immutable data with fields}
Since in ClassLess Java there are no classes, our \Q@Point2D@ must be an interface, as for example:
\begin{lstlisting}
interface Point2D{ int x(); int y();}
\end{lstlisting}
That is, since Java disallows fields inside interfaces,
 we simulate state by abstract methods.
By using the new Java8 \Q@default@ methods, we can easily add implemented methods inside the interface. For example:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  default float distanceFromOrigin(){ return Math.sqrt(x()*x()+y()*y());}
  }
\end{lstlisting}
resemble a class that have fields (getters) and methods.
However, how can we create an instance of \Q@Point2D@ without using classes? 
we can just provide an implementation for \Q@x()@ and \Q@y()@ as in
\begin{lstlisting}
new Point2D(){ public int x(){return 4;} public int y(){return 2;}}
\end{lstlisting}
However we are not required to use this cumbersome syntax\footnote{Available in Java from version ...} for every object allocation.
As programmes do, for ease or reuse, we can encapsulate boring long repetitive code in a method. A static factory method in this case is appropriate:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
  }
\end{lstlisting}
This obvious constructor code can be automatically generated by our \mixin annotation.

@@

More precisely, a field or factory parameter is generated for every no-args method requiring implementation whose name does not have special meaning\footnote{Formal definition of \mixin behaviour is provided later.}.
An example of code using \Q@Point2D@ is 
\Q@Point2D.of(42,myPoint.y())@
where we return a new point, using \Q@42@ as x-coordinate, and 
taking all the other informations (only \Q@y@ in this case) from another point.
This pattern is very common when programming with immutable data-structures; it is so common that we decided to support it in our code generation as \Q@with-@ methods, that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x();    int y();
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
is expanded by our annotation into

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
If the programmer would like a different implementation, they can just provide it using \Q@default@; that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}
is expanded into
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x;    int y=_y;
    public int x(){return x;}    public int y(){return y;}
    };}
  default Point2D withX(int val){ ... }    default Point2D withY(int val){ ... }}
\end{lstlisting}

That is, we can provide implementation for methods in an interface, 
and our annotation generate code only if the method \emph{needs} implementation. In this way is trivial for the programmer to personalize the behaviour if they have any special need.

@@

Just to give you a feeling on how much boring code \mixin is generating,  we show the expanded code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);    Point3D withY(int val);    Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int _x, int _y, int _z){
    int x=_x;    int y=_y;    int z=_z;
    return new Point3D(){
      public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
      public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
      public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
      public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
      public Point3D with(Point2D val){
        if(val instanceof Point3D){return (Point3D)val;}
        return Point3D.of(val.x(),val.y(),this.z());
        }      };    }  }
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

\Q@Bird@s are \Q@Animal@s, 
but while \Q@Animal@s, only need 2D locations,
\Q@Bird@s need  3D locations.
In Java if we define an animal class with a field we have a set of unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@:
this is bad since all animals would require more that is needed, and also it may requires the programmer to predict the future, or it may require to adapt the old code to accommodate for new evolutions.
\item Define a \Q@Point2D@ field in \Q@Animal@ and:
\item Define an extra \Q@int z@ field in \Q@Bird@.
This solution is very at-hoc, requires to basically duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside of \mbox{\Q@Bird@}.
Again, there are many reasons this would be bad, the most dramatic is that it would not scale to a scenario when the programmer of \Q@Bird@ and the programmer of \Q@Point3D@ are different.
\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects in the field and cast
the value out of the \Q@Point2D@ field to \Q@Point3D@ when implementing the overridden getter.
This solution scales to the multiple programmers approach, but requires ugly casts and can be implemented in a wrong way leading to bugs.
\end{itemize}
